---
title: "Driver Lifetime Value - Lyft"
author: "David Freifeld"
date: "Thursday, August 06, 2015"
output: html_document
---

## Synopsis

## Preprocessing the Data

I start by reading the data into R and doing some minor formatting. Then I convert the UNIX timestamps to actual dates, and order the rides chronologically within each driver id.

```{r warning=FALSE}
library(ggplot2)
library(plyr)
library(scales)

# first read in the data file 
setwd("C:/Users/David/workspace/lyft/")
lyftData <- read.csv('lyft_data_1.csv')
colnames(lyftData) <- c('id', 'donation', 'timestamp')
lyftData$id <- factor(lyftData$id)

# convert the UNIX timestamps (will assume PST because Lyft is in CA)
lyftData$timestamp <- as.POSIXct(lyftData$timestamp, 
    origin='1970-01-01', tz='US/Pacific')
lyftData$date <- as.Date(lyftData$timestamp, tz = 'US/Pacific')

# order each driver's data by timestamp
driverSplits <- split(lyftData, lyftData$id)
lyftDataList <- lapply(driverSplits, function(x) {
	x[order(x$timestamp),]
})
lyftData <- ldply(lyftDataList)
lyftData$.id <- NULL
```

## Exploratory Analysis

First let's explore the average donation amount received by each driver. I start by looking at the first 25 drivers only, in order to fit all of the boxplots on a single graph:

```{r}
ggplot(lyftData[lyftData$id %in% as.character(0:25),], aes(x = id, y = donation, fill = id)) +
    geom_boxplot() + 
    xlab("Driver ID") + 
    ylab("Donation Amount ($)") + 
    ggtitle("Distribution of Donation Amount by Driver") +
    theme(legend.position="none")
```

This plot shows that the distributions of donation amounts for each of the first 25 drivers are very similar. The median is $9 or $10 in every case. It turns out that this is true for all 98 drivers. (The plot for all 98 drivers looks identical to the above plot, and therefore I am leaving it out for visual purposes). 

Because the donation amount does not differ from driver to driver, we can use the average donation amount later in the calculation of driver lifetime value:

```{r}
avgDonation <- mean(lyftData$donation)
round(avgDonation, 2)
```

Next I look at the distributions of rides over time for each user. (Again, I am only using 25 drivers for illustrative purposes. The full plot of 98 drivers looks similar).

```{r warning=FALSE}
ggplot(lyftData[lyftData$id %in% as.character(26:50),], aes(x = timestamp)) +
	geom_histogram() +
	facet_wrap(~ id) + 
    ylab("Number of Rides") +
    xlab("Time") + 
    ggtitle("Distributions of Rides Over Time by Driver") + 
    scale_x_datetime(breaks = date_breaks("1 month"), labels=date_format("%m/%d"), 
                     limits=c(min(lyftData$timestamp), max(lyftData$timestamp)))
```



```{r}
# First get some summary statistics related to how long drivers
# keep driving
lastRideRecorded <- max(lyftData$date)

driverSummary <- ddply(lyftData, .(id), summarize,
	numRides = length(date),
	firstRide = date[1],
	lastRide = date[numRides],
	duration = as.numeric(lastRide - firstRide) + 1,
	stillActive = as.numeric(lastRideRecorded - lastRide) <= 7,
	uniqueDays = length(unique(date)),
	fractionWorked = uniqueDays / duration,
	ridesPerActiveDay = numRides/uniqueDays,
	maxBreak = as.numeric(max(date[2:numRides] - date[1:(numRides-1)])) - 1
)
rownames(driverSummary) <- driverSummary$id
activeDrivers <- driverSummary$stillActive
numStillActive <- sum(activeDrivers)
numDrivers <- nrow(driverSummary)
numDrivers

ggplot(driverSummary, aes(x = duration, y = fractionWorked)) + 
	geom_point(aes(fill = stillActive, size = ridesPerActiveDay), 
		position = 'jitter', shape = 21) +
	scale_fill_brewer(palette="Set1")

```

## Estimating Expected Driver Lifetime

```{r}
t <- 1:83
activeAtT <- sapply(t, function(x) {
    (sum(driverSummary[!activeDrivers, "duration"] > x) + numStillActive) / 
		numDrivers
})
survivalData <- data.frame(t = t, activeAtT = activeAtT)

ggplot(survivalData, aes(x = t, y = log(activeAtT))) + 
	geom_point() + 
	stat_smooth(method = "lm", formula = y ~ x - 1) + 
	ylim(-1, 0)

mod <- lm(log(activeAtT) ~ t - 1, data = survivalData)
summary(mod)

lambda <- -mod$coef[[1]]
expectedLife <- 1 / lambda
```

## Calculating a Driver Lifetime Value

```{r}
recordedRides <- nrow(lyftData)
recordedRides

aRidesPerActiveDay <- mean(driverSummary[activeDrivers, "ridesPerActiveDay"])
aFractionWorked <- mean(driverSummary[activeDrivers, "fractionWorked"])
futureRides <- aRidesPerActiveDay * aFractionWorked * expectedLife * numStillActive
futureRides

lyftRate <- 0.2

avgRides <- (recordedRides + futureRides) / numDrivers
avgRides

DLV <- avgRides * avgDonation * lyftRate
DLV
```

## Conclusion